---
title: "Combine Bench Over Time"
author: "Bernard Scott"
format: revealjs
editor: visual
message: false
---

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(ggthemes)
library(modeldata)
library(dplyr)
library(knitr)
library(kableExtra)
library(parsnip)
library(recipes)
library(tidymodels)
library(ggpubr)

theme_set(theme_minimal(base_size = 13))

combine <- read_csv("data/clean_combine.csv")

glimpse(combine)

condensed_combine <- combine %>%
  mutate(position = fct_collapse(position,
    "S" = c("FS", "SS"),
    "LB" = c("OLB", "ILB"),
    "DT" = c("DT", "NT")
  ))

glimpse(condensed_combine)
```

## Initial Research Question

- Each of the thirty-two National Football League (NFL) Franchises (teams) rosters a total of 53 players, with up to 16 on a "practice squad", and the remaining on the official team roster
- After every season, each team has the opportunity to draft players from a candidate pool
- Each player wanting to be drafted participates in the NFL Scouting Combine
- **I would like to know which biometrics of Combine participants are the best predictors of the number of bench press repetitions one can perform?**

## NFL Scouting Combine

![Player Sprinting the Forty Yard Dash at the NFL Scouting Combine, image from ("Washington's John Ross sets 40-yard dash record at NFL Scouting Combine" [https://www.wkyc.com/article/sports/nfl/nfl-draft/washingtons-john-ross-sets-40-yard-dash-record-at-nfl-scouting-combine/95-419799888](https://www.wkyc.com/article/sports/nfl/nfl-draft/washingtons-john-ross-sets-40-yard-dash-record-at-nfl-scouting-combine/95-419799888))](NFL_Combine_40_yard_dash_recent.jpg){#fig-forty_yard_dash width=70%}

## 225-lb Bench Press

![Official Bench Press bench used in the NFL Scouting Combine, image from ("NFL Combine: Which player did the most bench-press reps on Friday?" [https://www.al.com/sports/2019/03/nfl-combine-which-player-did-the-most-bench-press-reps-on-friday.html](https://www.al.com/sports/2019/03/nfl-combine-which-player-did-the-most-bench-press-reps-on-friday.html))](NFL_Combine_Bench_Front.avif){#fig-combine_bench_press width=70%}

## Combine Data

```{r}
#| message: false
#| label: tbl-condensed-combine
#| tbl-cap: "Select columns of the first five rows of the cleaned and condensed `combine` data set."

condensed_combine %>%
  select(name, position, heightinchestotal, weight, arms) %>%
  head(3) %>%
  kable()
```

## Player Bench Press Repetition Data

```{r}
#| label: fig-bench-over-time
#| fig-cap: "Exploratory plot showing the average number of bench press repetitions from 1999 to 2015."

condensed_combine %>%
  group_by(year) %>%
  summarise(mean_bench = mean(bench, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = mean_bench)) +
  geom_point() +
  scale_x_continuous(breaks = 1999:2015) +
  geom_smooth(method=lm, se=FALSE) +
  theme(axis.text.x = element_text(angle=45)) + 
  labs(x = "Year", y = "Average Bench Press Reps", title = "Average Bench Press Reps by Year")
```

## Player Weight Data

```{r}
#| label: fig-weight-hist
#| fig-cap: "Histogram of combine participant weight values."

condensed_combine %>%
  ggplot(aes(x = weight)) +
  geom_histogram(bins=60) +
  labs(x = "Weight (lbs)", y = "Number of Players")
```

## Player Height Data

```{r}
#| label: fig-height-hist
#| fig-cap: "Histogram of combine participant heights."

condensed_combine %>%
  ggplot(aes(x = heightinchestotal)) +
  geom_histogram(bins=40)+
  labs(x = "Height (in)", y = "Number of Players")
```

## Player Arm Length Data

```{r}
#| label: fig-arm-length-hist
#| fig-cap: "Histogram of combine participant arm lengths."

condensed_combine %>%
  filter(arms != 0) %>%
  ggplot(aes(x = arms)) +
  geom_histogram(bins = 60) +
  labs(x = "Arm Length (in)", y = "Number of Players")
```

## Player Weight and Number of Repetitions

```{r}
#| label: fig-bench-vs-weight-over-time
#| fig-cap: "Plot showing relationship between number of bench press repetitions and weight, one plot for each year included in the data set. Weight is in units of pounds (lbs)."

bench_weight <- function(yr) {
  condensed_combine %>%
    filter(bench > 0 & year == yr) %>%
    ggplot(aes(x = bench, y = weight)) +
    geom_point() +
    geom_smooth(method=lm, se=FALSE) +
    theme(axis.text.y=element_blank()) +
    labs(x = "Bench Reps", y = "Weight")
}

bench_weight_per_year <- map(1999:2015, bench_weight)

ggarrange(plotlist = bench_weight_per_year, ncol = 4, nrow = 5)
```

## Player Height and Number of Repetitions

```{r}
#| label: fig-bench-vs-height-over-time
#| fig-cap: "Plot showing relationship between number of bench press repetitions and height, one plot for each year included in the data set. Height is in units of inches (in)."

bench_height <- function(yr) {
  condensed_combine %>%
    filter(bench > 0 & year == yr & heightinchestotal > 0) %>%
    ggplot(aes(x = bench, y = heightinchestotal)) +
    geom_point() +
    geom_smooth(method=lm, se=FALSE) +
    theme(axis.text.y=element_blank()) +
    labs(x = "Bench Reps", y = "Height")
}

bench_height_per_year <- map(1999:2015, bench_height)

ggarrange(plotlist = bench_height_per_year, ncol = 4, nrow = 5)
```

## Player Arm Length and Number of Repetitions

```{r}
#| label: fig-bench-vs-arm-length-over-time
#| fig-cap: "Plot showing relationship between number of bench press repetitions and height, one plot for each year included in the data set. Arm Length is in units of inches (in)."

bench_arms <- function(yr) {
  condensed_combine %>%
    filter(bench > 0 & year == yr & position != "K" & arms != 0) %>%
    ggplot(aes(x = bench, y = arms)) +
    geom_point() +
    geom_smooth(method=lm, se=FALSE) +
    theme(axis.text.y=element_blank()) +
    labs(x = "Bench Reps", y = "Arm Length")
}

bench_height_per_year <- map(1999:2015, bench_arms)

ggarrange(plotlist = bench_height_per_year, ncol = 4, nrow = 5)
```

## Model Estimates for Number of Repetitions

```{r}
#| label: fig-lin-reg
#| fig-cap: "Linear regression predicting the number of repetitions of bench press a player can perform based on their height, weight, and arm length."

# set seed for reproducability
set.seed(4133)

# load the combine dataset (should already be loaded from previous chunks)
combine_nums <- condensed_combine %>%
  select(bench, arms, heightinchestotal, weight) %>%
  filter(bench != 0, arms != 0, heightinchestotal != 0, weight != 0)

#data(combine_nums)

# create a training and testing split
combine_split <- initial_split(combine_nums, prop = 0.8)

# create a training and testing dataset
combine_training <- training(combine_split)
combine_testing <- testing(combine_split)

# create a recipe for preprocessing
combine_recipe <- recipe(bench ~ ., data = combine_training) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

# create a linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

# fit the model
lm_fit <- workflow() %>%
  add_recipe(combine_recipe) %>%
  add_model(lm_spec) %>%
  fit(data = combine_training)

# make predictions
lm_pred <- predict(lm_fit, combine_testing) %>%
  bind_cols(combine_testing)

lm_pred %>%
  ggplot(aes(x = bench, y = .pred)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Bench Press Repetitions", y = "Predicted Number of Repetitions")
```

## Bench Press Repetition Predictions

```{r}
#| label: fig-bench-press-lin-reg
#| fig-cap: The results of a linear regression predicting the number of repetitions of bench press a player can perform based on the player's height, weight, and arm length.

# evaluate the model
lm_eval <- lm_pred %>%
  metrics(truth = bench, estimate = .pred) %>%
  bind_rows(
    lm_pred %>%
      rsq(truth = bench, estimate = .pred) %>%
      mutate(metric = "rsq")
  )

# tidy the model results
lm_tidy <- tidy(lm_fit)

# visualize the model results
lm_tidy %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) +
  coord_flip() +
  labs(
    title = "Linear Regression Model Results",
    subtitle = "Model: bench ~ .",
    x = "Predictor",
    y = "Estimate"
  )
```

## Model Estimates for Player Position

```{r}
#| label: fig-knn-classification-results
#| fig-cap: A K-Nearest-Neighbors classification categorizing players by which group of positions they play on the field, as an offensive/defensive lineman or secondary player.

generalized_combine <- condensed_combine %>%
  mutate(position = fct_collapse(position,
    "O-Line" = c("OG", "OT", "C"),
    "O-2ry" = c("FB", "QB", "RB", "TE", "WR"),
    "D-Line" = c("DT", "DE"),
    "D-2ry" = c("CB", "S", "LB"),
    "Special" = c("K", "P", "LS"),
  ))

#glimpse(generalized_combine)

# set seed for reproducability
set.seed(4133)

# load the combine dataset (should already be loaded from previous chunks)
combine_pos <- generalized_combine %>%
  select(heightinchestotal, weight, arms, hands, fortyyd, twentyyd, tenyd, twentyss, threecone, vertical, broad, bench, position) #%>%
  #filter(weight != 0, arms != 0, hands != 0, fortyyd != 0, twentyyd != 0, tenyd != 0, twentyss != 0, threecone != 0, vertical != 0, broad != 0, bench != 0)

# glimpse(combine_pos)

# create a training and testing split
combine_pos_split <- initial_split(combine_pos, prop = 0.8)

# create a training and testing dataset
combine_pos_training <- training(combine_pos_split)
combine_pos_testing <- testing(combine_pos_split)

# every predictor adds a dimension, and I have too many of them.

position_rec <- recipe(position ~ heightinchestotal + weight + arms + hands + fortyyd + twentyyd + tenyd + twentyss + threecone + vertical + broad + bench,
                       data = combine_pos_training) %>%
  step_normalize(all_numeric_predictors())

position_spec <- nearest_neighbor(neighbors = 5) %>%
  set_engine("kknn") %>%
  set_mode("classification")

#neighbors = tune()

#follow lesson 8 example 3

position_wf <- workflow() %>%
  add_recipe(position_rec) %>%
  add_model(position_spec)

position_fit <- fit(position_wf,combine_pos_training)

#| label: fig-knn
#| fig-cap: "K-Nearest-Neighbors regression categorizing players by position based upon their numerical combine statistics."

position_fit %>%
  predict(combine_pos_testing) %>%
  cbind(combine_pos_testing) %>%
  mutate(prediction = ifelse(position == .pred_class, "Correct", "Incorrect")) %>%
  ggplot(aes(x = fortyyd, y = weight, color = prediction)) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_colorblind() +
  labs(x = "Weight (lb)", y = "Forty Yard Dash (yd)")
```

## Player Position Model Accuracy

```{r}
#| label: tbl-knn-accuracy
#| tbl-cap: "Accuracy of the K-Nearest-Neighbors model, displayed in the `.estimate` column."
position_fit %>%
  predict(combine_pos_testing) %>%
  cbind(combine_pos_testing) %>%
  accuracy(position,.pred_class) %>%
  kable()
```

## Player Position Model Performance

```{r}
#| label: fig-knn-truth
#| fig-cap: "Truth table of the K-Nearest-Neighbors classification where the values on the diagonals are successful identifications, and any others are unsuccessful identifications."
position_fit %>%
  predict(combine_pos_testing) %>%
  cbind(combine_pos_testing) %>%
  conf_mat(position,.pred_class) %>%
  autoplot(type = "heatmap")
```

## Conclusions

- Evidence supporting the hypothesis that player weight, height, and arm length has an impact on NFL Scouting Combine bench press performance
- Individual player position is not likely to have an impact on their bench press performance
- Lack of evidence that player position can be determined solely by observing Combine statistics

## References

- [NFLSavant.com](https://nflsavant.com)
- [The National Football League](https://en.wikipedia.org/wiki/National_Football_League)
- [NFL Scouting Combine](https://en.wikipedia.org/wiki/NFL_Scouting_Combine)
- [Evolution of the Player](https://operations.nfl.com/inside-football-ops/players-legends/evolution-of-the-nfl-player/#:~:text=Over%20time%2C%20teams'%20roster%20sizes,inactive%20players%20before%20each%20game.)
- [Diagram highlighting muscles worked when performing a bench press](https://healthybodyathome.com/how-and-why-you-should-bench-press/)